Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-06-02 aosd13/paperBody.tex
Misc
Chap. 5: Revisions

\chapter{Introduction}
In-memory collections of data often need efficient processing. For on-disk
data, efficient processing is already provided by database management systems~(DBMS) thanks to
their query optimizers, which support many optimizations specific to the domain
of collections. Moving in-memory data to DBMSs, however,
typically does not improve performance~\citep{Stonebraker07}, and query optimizers cannot
be reused separately since DBMS are typically monolithic and their optimizers deeply integrated.
A few collection-specific optimizations, such as shortcut
fusion~\citep{Gill93shortcut}, are supported by compilers for purely
functional languages such as Haskell. However, the implementation techniques for
those optimizations do not
generalize to many other ones, such as support for indexes. In
general, collection-specific optimizations are not supported by the general-purpose
optimizers used by typical (JIT) compilers.

@@

For this reason, our approach is implemented as an embedded domain-specific language, that is, as a regular library.
We call this library \LoSDef. \LoS\ consists of a domain-specific language (DSL) for queries on collections based on the Scala collections API\@. 
This DSL is implemented as an embedded DSL (EDSL) for Scala. An expression in this EDSL produces at run time an \emph{expression tree} in the host language: a data structure which represents the query to execute, similar to an abstract syntax tree (AST) or a query plan. Thanks to the extensibility of Scala, expressions in this language look almost identical to expressions with the same meaning in Scala.
When executing the query, \LoS\ optimizes and compiles these expression trees for more efficient execution. Doing optimization at run time, instead of compile-time, avoids the need for control-flow analyses to determine which code will be actually executed~\citep{Chambers10}, as we will see later.

We have choosen Scala \citep{Odersky11book} to implement our library for two reasons: (i) Scala is a good meta-language for embedded DSLs, because it is syntactically
flexible and has a powerful type system, and (ii) Scala has a sophisticated collections library with an attractive syntax (for-comprehensions) to specify queries.

>>>>

\chapter{Introduction}
\label{sec:aosd13-intro}
In-memory collections of data often need efficient processing. For on-disk
data, efficient processing is already provided by database management systems~(DBMS) thanks to
their query optimizers, which support many optimizations specific to the domain
of collections. Moving in-memory data to DBMSs, however,
typically does not improve performance~\citep{Stonebraker07}, and query optimizers cannot
be reused separately since DBMS are typically monolithic and their optimizers deeply integrated.
A few collection-specific optimizations, such as shortcut
fusion~\citep{Gill93shortcut}, are supported by compilers for purely
functional languages such as Haskell. However, the implementation techniques for
those optimizations do not
generalize to many other ones, such as support for indexes. In
general, collection-specific optimizations are not supported by the general-purpose
optimizers used by typical (JIT) compilers.

@@

For this reason, our approach is implemented as an embedded domain-specific language (EDSL), that is, as a regular library.
We call this library \LoSDef. \LoS\ consists of a Scala EDSL for queries on collections based on the Scala collections API\@.
An expression in this EDSL produces at run time an \emph{expression tree} in the host language: a data structure which represents the query to execute, similar to an abstract syntax tree (AST) or a query plan. Thanks to the extensibility of Scala, expressions in this language look almost identical to expressions with the same meaning in Scala.
When executing the query, \LoS\ optimizes and compiles these expression trees for more efficient execution. Doing optimization at run time, instead of compile-time, avoids the need for control-flow analyses to determine which code will be actually executed~\citep{Chambers10}, as we will see later.

We have chosen Scala \citep{Odersky11book} to implement our library for two reasons: (i) Scala is a good meta-language for EDSLs, because it is syntactically
flexible and has a powerful type system, and (ii) Scala has a sophisticated collections library with an attractive syntax (for-comprehensions) to specify queries.

### 2017-06-02 aosd13/pepm13/caseStudyImpl.tex
Chap. 5: Revisions

However, given the design of the collection library, every instance of \code{TraversableLike[T, Repr]} is also an instance of both \code{Traversable[T]} and \code{Repr}; to take advantage of this during interpretation and optimization, we need to restrict the type of \code{expToTraversableLikeOps}, getting the following conversion:\footnote{Due to type inference bugs, the actual implicit conversion needs to be slightly more complex, to mention \code{T} directly in the argument type. We reported the bug at \url{https://issues.scala-lang.org/browse/SI-5298}.}

>>>>

Subclasses of \code{TraversableLike[T, Repr]} also subclass both \code{Traversable[T]} and \code{Repr}; to take advantage of this during interpretation and optimization, we need to restrict the type of \code{expToTraversableLikeOps}, getting the following conversion:\footnote{Due to type inference bugs, the actual implicit conversion needs to be slightly more complex, to mention \code{T} directly in the argument type. We reported the bug at \url{https://issues.scala-lang.org/browse/SI-5298}.}

### 2017-06-02 aosd13/pepm13/caseStudyIntro.tex
Chap. 5: Revisions

Such optimizations are not possible within most ordinary collections API, such as the Scala collections API~\cite{odersky2009fighting}, because they
do not reify queries on collections. Instead, each API function is defined compositionally in terms
of the meaning of the other API functions.
We regard the Scala collections API as a \emph{domain-specific embedded language} (EDSL); this style of compositional DSL implementation by libraries is often
called \emph{shallow embedding}%
%~\chk{add reference} %PG: I know it's crazy, but nobody gives a reference for this term.
. The alternative, which we pursue, is a \emph{deep embedding} of the DSL, where programs written in the DSL (in our case: 
queries on collections) are reified as data or, more specifically, as abstract syntax trees, which can be subject to non-compositional semantics. Among others, they can then be analyzed, optimized, transformed, interpreted, or compiled.

>>>>

Such optimizations are not possible within most ordinary collections API, such as the Scala collections API~\cite{odersky2009fighting}, because they
do not reify queries on collections. Instead, each API function is defined compositionally in terms
of the meaning of the other API functions.
We regard the Scala collections API as a \emph{embedded domain-specific language} (EDSL); this style of compositional DSL implementation by libraries is often
called \emph{shallow embedding}%
%~\chk{add reference} %PG: I know it's crazy, but nobody gives a reference for this term.
. The alternative, which we pursue, is a \emph{deep embedding} of the DSL, where programs written in the DSL (in our case: 
queries on collections) are reified as data or, more specifically, as abstract syntax trees, which can be subject to non-compositional semantics. Among others, they can then be analyzed, optimized, transformed, interpreted, or compiled.
